Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-06 submission.tex
Moving around, fixes to sec:setup
added transition from sec:prev to sec:data

\section{Introduction}
\label{sec:intro}
Part-of-speech (PoS) tagging is an important preprocessing step for many NLP
tasks, such as dependency parsing \cite{Niv:Hal:Kub:07,Haj:Cia:Joh:09}, named
entity recognition \cite{Tjo:DeM:03} and sentiment analysis
\cite{Wil:Wie:Hof:09}. Whereas much effort has gone into the development of
PoS-taggers -- to the effect that this task is often considered more or less a
solved task -- considerably less effort has been devoted to the empirical
evaluation of the PoS tag sets themselves. Error analysis of PoS taggers
indicate that, whereas tagging improvement through means of learning algorithm
or feature engineering seems to have reached something of a plateau, linguistic
and empirical assessment of the distinctions made in the PoS tag sets may be a
venue worth investigating further \cite{Man:11}. Clearly, the utility of a PoS
tag set is tightly coupled with the downstream task for which it is performed.
Even so, PoS tag sets are usually employed in a ``one size fits all'' fashion,
regardless of the requirements posed by the task which makes use of this
information.

@@

% fixme:
%Transition/overgang her ...

@@

The syntactic annotation choices in NDT are largely based on the Norwegian
Reference Grammar \cite{Faa:Lie:Van:97}. Some central annotation choices are outlined in
Table \ref{ndtannotation}, taken from of \newcite{Sol:Skj:Ovr:14},
providing overview of the analyses of syntactic constructions that often
distinguish dependency treebanks, such as coordination and the treatment of
auxiliary and main verbs. The annotations comprise 29
dependency relations, including \textsc{adv} (adverbial), \textsc{subj}
(subject) and \textsc{koord} (coordination).

@@

As we want to investigate how we can improve the linguistic quality of a tag
set, we will only consider tag set modifications we deem linguistically
sensible.

@@

%This is done by concatenating the original tag with the
%specified features for all applicable tokens

@@

\paragraph{Baseline}
As a point of reference for the PoS tagging we will be reporting the commonly used
most-frequent-tag baseline, henceforth MFT.
This involves labeling each word with the tag it was
assigned most frequently in the training data. All unknown words, i.e., words not
seen in the training data, are assigned the tag most frequently assigned to
words seen only once.  The rationale is that we might expect unknown
and infrequent words to have similar properties given that they both occur rarely.

@@

Table \ref{verbresults} presents the results from tagging and parsing with
modified verb tags. Introducing features in isolation, mood is the only feature
leading to an increase in LAS, with a reported LAS of 87.04\%. Imperative
clauses are fundamentally different from indicative clauses, as they lack an
overt subject.  Although voice is a very distinguishing property for verbs, and
passive verbs appear in clauses very different from those with active verbs,
introducing this distinction in the tag set leads to a drop in LAS of 0.05
percentage points.  Distinguishing between the various tenses yields an LAS of
86.97\%. Combining the two most promising features, mood and tense, resulted in
an LAS of 87.12\% and UAS of 90.31\%.

@@

The results from the experiments with determiners are shown in Table
\ref{detresults}.  Introducing the distinction of type led to an increase in
tagger accuracy by 0.14 percentage points to 97.61\%, while marginally
impacting the parsing, with LAS of 87.00\%, 0.01 percentage points below that
of the original tag set.  The increase in tagger accuracy when introducing the
distinction of type is noteworthy, as we expected the finer granularity to lead
to a decrease in accuracy. This serves to indicate that more fine-grained
distinctions for determiners, which is a quite disparate category in the
treebank, may be quite useful for tagging. However, as it has negative impact
on the syntactic parsing, we can conclude that the type of a determiner does
not assist in generalizing syntactic patterns, as most determiners, regardless
of type, appear in the same syntactic constructions (i.e., before a noun or
noun phrase).

>>>>

\section{Introduction}
\label{sec:intro}
Part-of-speech (PoS) tagging is an important preprocessing step for many NLP
tasks, such as dependency parsing \cite{Niv:Hal:Kub:07,Haj:Cia:Joh:09}, named
entity recognition \cite{Tjo:DeM:03} and sentiment analysis
\cite{Wil:Wie:Hof:09}. Whereas much effort has gone into the development of
PoS taggers -- to the effect that this task is often considered more or less a
solved task -- considerably less effort has been devoted to the empirical
evaluation of the PoS tag sets themselves. Error analysis of PoS taggers
indicate that, whereas tagging improvement through means of learning algorithm
or feature engineering seems to have reached something of a plateau, linguistic
and empirical assessment of the distinctions made in the PoS tag sets may be a
venue worth investigating further \cite{Man:11}. Clearly, the utility of a PoS
tag set is tightly coupled with the downstream task for which it is performed.
Even so, PoS tag sets are usually employed in a ``one size fits all'' fashion,
regardless of the requirements posed by the task which makes use of this
information.

@@

We argue, however, that it is possible to significantly improve the performance
of the downstream application of syntactic parsing by introducing
linguistically informed, syntactically informative distinctions in a tag set.
We will now turn to the data that provided the basis for our experiments and
outline its morphological and syntactic annotation choices.

@@

The syntactic annotation choices in NDT are largely based on the Norwegian
Reference Grammar \cite{Faa:Lie:Van:97}. Some central annotation choices are
outlined in Table \ref{ndtannotation}, taken from \newcite{Sol:Skj:Ovr:14},
providing overview of the analyses of syntactic constructions that often
distinguish dependency treebanks, such as coordination and the treatment of
auxiliary and main verbs. The annotations comprise 29 dependency relations,
including \textsc{adv} (adverbial), \textsc{subj} (subject) and \textsc{koord}
(coordination).

@@

@@

As we want to investigate how we can improve the linguistic quality of a tag
set, we will only consider tag set modifications we deem linguistically
sensible.
%This is done by concatenating the original tag with the
%specified features for all applicable tokens

@@

\paragraph{Baseline}
As a point of reference for the PoS tagging, we will be reporting the commonly
used most-frequent-tag baseline, henceforth MFT. This involves labeling each
word with the tag it was assigned most frequently in the training data. All
unknown words, i.e., words not seen in the training data, are assigned the tag
most frequently assigned to words seen only once. The rationale is that we
might expect unknown and infrequent words to have similar properties given that
they both occur rarely.

@@

Table \ref{verbresults} presents the results from tagging and parsing with
modified verb tags. Imperative clauses are fundamentally different from
indicative clauses, as they lack an overt subject, which is reflected in the
fact that mood is the only feature leading to an increase in LAS, with a
reported LAS of 87.04\% Although voice is a very distinguishing property for
verbs, and passive verbs appear in clauses very different from those with
active verbs, introducing this distinction in the tag set leads to a drop in
LAS of 0.05 percentage points.  Distinguishing between the various tenses
yields an LAS of 86.97\%. Combining the two most promising features, mood and
tense, resulted in an LAS of 87.12\% and UAS of 90.31\%.

@@

The results from the experiments with determiners are shown in Table
\ref{detresults}.  Introducing the distinction of type led to an increase in
tagger accuracy by 0.14 percentage points to 97.61\%, while marginally
impacting the parsing, with LAS of 87.00\%, 0.01 percentage points below that
of the original tag set.  The increase in tagger accuracy when introducing the
distinction of type is noteworthy, as we expected the finer granularity to lead
to a decrease in accuracy. This serves to indicate that more fine-grained
distinctions for determiners, which is a quite disparate category in the
treebank, may be quite useful for tagging. However, as it has negative impact
on parsing, we can conclude that the type of a determiner does not assist in
generalizing syntactic patterns. Most determiners, regardless of type, appear
in the same syntactic constructions (i.e., before a noun or noun phrase).
